---
title: "R Notebook"
output: html_notebook
---

Dit is een R Notebook. Deze gaat ervanuit dat je beschikt over de R Notebook 'po_analyse.Rmd'. 


```{r}
school_herverdeling_data = baos_2019 %>% 
  mutate(
    diff_nl = ll_lr_ratio - (sum(leerlingen)/ sum(onderwijzers)),
    leraar_overschot_tekort = onderwijzers - (leerlingen / (sum(leerlingen) / sum(onderwijzers))),
    perc_leraar_overschot_tekort = (leraar_overschot_tekort / onderwijzers) * 100
  )
  
library(ggplot2)  

ggplot(school_herverdeling_data, aes(gem_loonsom, leraar_overschot_tekort)) +
  geom_bin2d()

ggplot(school_herverdeling_data, aes(gem_loonsom, perc_leraar_overschot_tekort)) +
  geom_bin2d()

ggplot(school_herverdeling_data, aes(gem_loonsom, ll_lr_ratio)) +
  theme_minimal() + labs(fill = "Aantal scholen") + geom_bin2d(bins=100) +
  labs(x = 'Gemiddelde Loonsom', y = 'LLR')

summary(lm(data=school_herverdeling_data, ll_lr_ratio~gem_loonsom))


summary(lm(data=school_herverdeling_data, leraar_overschot_tekort~gem_loonsom))
summary(lm(data=school_herverdeling_data, perc_leraar_overschot_tekort~gem_loonsom))
  
school_herverdeling_data_gemeente = baos_2019 %>% 
  group_by(statcode) %>% mutate(
    diff_nl = ll_lr_ratio - (sum(leerlingen)/ sum(onderwijzers)),
    leraar_overschot_tekort = onderwijzers - (leerlingen / (sum(leerlingen) / sum(onderwijzers))),
    perc_leraar_overschot_tekort = (leraar_overschot_tekort / onderwijzers) * 100
  )

ggplot(school_herverdeling_data_gemeente, aes(gem_loonsom, leraar_overschot_tekort)) +
  geom_bin2d()

ggplot(school_herverdeling_data_gemeente, aes(gem_loonsom, perc_leraar_overschot_tekort)) +
  geom_bin2d()

summary(lm(data=school_herverdeling_data_gemeente, leraar_overschot_tekort~gem_loonsom))
summary(lm(data=school_herverdeling_data_gemeente, perc_leraar_overschot_tekort~gem_loonsom))


```
Geografische patronen in loonsom
```{r}
gemeente_lonen_2019 = baos_2019 %>% 
  group_by(statcode) %>% summarise(
    totaal_loonsom = sum(loonsom), 
    totaal_onderwijzers = sum(onderwijzers),
    gemiddeld_loonsom = totaal_loonsom / totaal_onderwijzers,
    )



baos_2014 = baos %>% filter(jaar==2014)

gemeente_lonen_2014 = baos_2014 %>% 
  group_by(statcode) %>% mutate(
    totaal_loonsom = sum(loonsom), 
    totaal_onderwijzers = sum(onderwijzers),
    gemiddeld_loonsom = totaal_loonsom / totaal_onderwijzers,
    )


  
```

Aanpassing dataset nodig nog: 


```{r}
library(sf)
gemeentegrenzen <- st_read("https://geodata.nationaalgeoregister.nl/cbsgebiedsindelingen/wfs?request=GetFeature&service=WFS&version=2.0.0&typeName=cbs_gemeente_2019_gegeneraliseerd&outputFormat=json")

data <-
  gemeentegrenzen %>%
  left_join(gemeente_lonen_2019)

data %>%
  ggplot() +
  geom_sf(aes(fill = gemiddeld_loonsom)) +
  scale_fill_viridis_c(limits=c(2700, 4000)) +
  labs(title = "Gemiddelde Loonsom 2019 (per gemeente)", fill = "") +
  theme_void()

data <-
  gemeentegrenzen %>%
  left_join(gemeente_lonen_2014)

data %>%
  ggplot() +
  geom_sf(aes(fill = gemiddeld_loonsom)) +
  scale_fill_viridis_c(limits=c(2700, 4000)) +
  labs(title = "Gemiddelde Loonsom 2014 (per gemeente)", fill = "") +
  theme_void()

```


```{r}
gemeente_loon_jaar = baos %>% 
  group_by(jaar) %>% mutate(
    jaar_gemiddelde = mean(loonsom)
  ) %>% select(jaar_gemiddelde, jaar)

ggplot(gemeente_loon_jaar, aes(x = jaar, y = jaar_gemiddelde)) + geom_line() + theme_minimal() 


ontwikkeling_loon_gemeente = baos %>%  filter(jaar == 2014 | jaar == 2019) %>% group_by(statcode, brin)  %>%
   summarise(
     dif_loon_school = gem_loonsom[jaar==2019] - gem_loonsom[jaar==2014] 
   ) %>% group_by(statcode) %>%
  summarise(
    gem_gemeente = mean(dif_loon_school)
  )

data <-
  gemeentegrenzen %>%
  left_join(ontwikkeling_loon_gemeente)

data %>%
  ggplot() +
  geom_sf(aes(fill = gem_gemeente)) +
  scale_fill_viridis_c() +
  labs(title = "Verandering Loonsom 2014-2019 (per gemeente)", fill = "") +
  theme_void()

```
```{r}
hist(baos_2019$gem_loonsom)

schaal_11_data = baos_2019 %>% mutate(
  percentage_11 = schaal_11 / onderwijzers
) 

ggplot(schaal_11_data, aes(percentage_11, gem_loonsom)) +
  geom_bin2d()

schaal_11_12_data = baos_2019 %>% mutate(
  percentage_11_12 = (schaal_11 + schaal_12) / onderwijzers
) 

ggplot(schaal_11_12_data, aes(percentage_11_12, gem_loonsom)) +
  geom_bin2d()

```

```{r}
baos_2019_schoolgewicht = baos_2019  %>% mutate(
   percentage_11_12 = (schaal_11 + schaal_12) / onderwijzers
)

ggplot(baos_2019, aes(gem_loonsom, schoolweging)) +
  geom_bin2d(bins=100) + theme_minimal() + labs(fill = "Aantal scholen") +
  labs(x = 'Gemiddelde Loonsom', y = 'Schoolgewicht')



summary(lm(data=baos_2019_schoolgewicht, gem_loonsom~schoolweging+percentage_11_12))


```

Vermogen
```{r}

baos_2019_besturen <- baos_2019 %>% 
  mutate(
     diff_nl = ll_lr_ratio - (sum(leerlingen)/ sum(onderwijzers)),
    leraar_overschot_tekort = onderwijzers - (leerlingen / (sum(leerlingen) / sum(onderwijzers)))
  ) %>% group_by(bevoegd_gezag) %>% summarise(
    per_leraar_tekort = (sum(leraar_overschot_tekort) / sum(onderwijzers)) *100,
    leraar_overschot_tekort_bestuur = sum(leraar_overschot_tekort), 
    gem_loonsom = mean(gem_loonsom),
    vermogen_bestuur = mean(vermogen),
    vermogen_leerling = sum(vermogen)/sum(leerlingen)
  ) 

require(scales)
ggplot(baos_2019_besturen %>% filter(vermogen_leerling < 150000), aes(gem_loonsom, vermogen_leerling)) +
  geom_bin2d(bins=100) + theme_minimal() + labs(fill = "Aantal scholen") +
  labs(x = 'Gemiddelde Loonsom', y = 'Vermogen per Leerling') + scale_y_continuous(labels = comma)

ggplot(baos_2019, aes(gem_loonsom, oab / leerlingen)) +
  geom_bin2d(bins=100) + theme_minimal() + labs(fill = "Aantal scholen") +
  labs(x = 'Gemiddelde Loonsom', y = 'Achterstandsmiddelen per leerling') + scale_y_continuous(labels = comma)

ggplot(baos_2019_besturen, aes(vermogen_bestuur, leraar_overschot_tekort_bestuur)) +
  geom_bin2d()

ggplot(baos_2019_besturen, aes(vermogen_bestuur, totaal_loonsom)) +
  geom_bin2d()

summary(lm(data=baos_2019_besturen, per_leraar_tekort~vermogen_bestuur))

ggplot(baos_2019_besturen, aes(, per_leraar_tekort)) +
  geom_bin2d()


```

```{r}
schaal_data = baos_2019 %>% mutate(
  percentage_11 = schaal_11 / onderwijzers,
  percentage_11_12 = (schaal_11 + schaal_12)/ onderwijzers
) 

ggplot(baos_2019, aes(gem_loonsom, leeftijd )) +
  geom_bin2d(bins=100) + theme_minimal() + labs(fill = "Aantal scholen") +
  labs(x = 'Gemiddelde Loonsom', y = 'Gemiddelde Leeftijd')

ggplot(schaal_data, aes(gem_loonsom, percentage_11_12 * 100)) +
  geom_bin2d(bins=100) + theme_minimal() + labs(fill = "Aantal scholen") +
  labs(x = 'Gemiddelde Loonsom', y = 'Percentage Schaal 11 of 12')

ggplot(baos_2019_schoolgewicht, aes(gem_loonsom, schoolweging)) +
  geom_bin2d(bins=100) + theme_minimal() + labs(fill = "Aantal scholen") +
  labs(x = 'Gemiddelde Loonsom', y = 'Schoolgewicht')

summary(lm(data=schaal_data, gem_loonsom~leeftijd))
summary(lm(data=schaal_data, gem_loonsom~leeftijd+percentage_11_12))
summary(lm(data=schaal_data, gem_loonsom~leeftijd+percentage_11))

ggplot(baos_2019, aes((totaal_bekostiging / leerlingen), gem_loonsom)) +
  geom_bin2d() + xlim(0,15000)

summary(lm(data=baos_2019 %>% mutate(totaal_leerling = totaal_bekostiging / leerlingen), gem_loonsom~totaal_leerling))
summary(lm(data=schaal_data %>% mutate(totaal_leerling = totaal_bekostiging / leerlingen), gem_loonsom~totaal_leerling+leeftijd+percentage_11_12))

summary(lm(data=baos_2019, gem_loonsom~oab))
summary(lm(data=schaal_data, gem_loonsom~oab+leeftijd+percentage_11_12))

salaris_model <- lm(data=schaal_data, gem_loonsom~leeftijd+percentage_11_12)

schaal_data$gem_loonsom_verschil = schaal_data$gem_loonsom - predict(salaris_model, schaal_data)

hist(schaal_data$gem_loonsom_verschil)

summary(lm(data=schaal_data %>% mutate(vermogen_ll = vermogen / leerlingen), gem_loonsom_verschil~vermogen_ll))



```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

