---
title: "R Notebook"
output: html_notebook
---

Dit is een R Notebook. Deze gaat ervanuit dat je beschikt over de R Notebook 'po_analyse.Rmd'. 


```{r}
school_herverdeling_data = baos_2019 %>% 
  mutate(
    diff_nl = ll_lr_ratio - (sum(leerlingen)/ sum(onderwijzers)),
    leraar_overschot_tekort = onderwijzers - (leerlingen / (sum(leerlingen) / sum(onderwijzers))),
    perc_leraar_overschot_tekort = (leraar_overschot_tekort / onderwijzers) * 100
  )
  
library(ggplot2)  

ggplot(school_herverdeling_data, aes(gem_loonsom, leraar_overschot_tekort)) +
  geom_bin2d()

ggplot(school_herverdeling_data, aes(gem_loonsom, perc_leraar_overschot_tekort)) +
  geom_bin2d()

summary(lm(data=school_herverdeling_data, leraar_overschot_tekort~gem_loonsom))
summary(lm(data=school_herverdeling_data, perc_leraar_overschot_tekort~gem_loonsom))
  
school_herverdeling_data_gemeente = baos_2019 %>% 
  group_by(statcode) %>% mutate(
    diff_nl = ll_lr_ratio - (sum(leerlingen)/ sum(onderwijzers)),
    leraar_overschot_tekort = onderwijzers - (leerlingen / (sum(leerlingen) / sum(onderwijzers))),
    perc_leraar_overschot_tekort = (leraar_overschot_tekort / onderwijzers) * 100
  )

ggplot(school_herverdeling_data_gemeente, aes(gem_loonsom, leraar_overschot_tekort)) +
  geom_bin2d()

ggplot(school_herverdeling_data_gemeente, aes(gem_loonsom, perc_leraar_overschot_tekort)) +
  geom_bin2d()

summary(lm(data=school_herverdeling_data_gemeente, leraar_overschot_tekort~gem_loonsom))
summary(lm(data=school_herverdeling_data_gemeente, perc_leraar_overschot_tekort~gem_loonsom))


```
Geografische patronen in loonsom
```{r}
gemeente_lonen_2019 = baos_2019 %>% 
  group_by(statcode) %>% mutate(
    totaal_loonsom = sum(loonsom), 
    totaal_onderwijzers = sum(onderwijzers),
    gemiddeld_loonsom = totaal_loonsom / totaal_onderwijzers,
    ) %>% select(
      statcode, totaal_loonsom, totaal_onderwijzers, gemiddeld_loonsom, loc_randstad
    )

summary(lm(data=gemeente_lonen_2019, gemiddeld_loonsom~loc_randstad))

boxplot(gemiddeld_loonsom~loc_randstad, data=gemeente_lonen_2019, xlab= "Randstad", ylab="Gemiddeld Loon")
hist(gemeente_lonen_2019$gemiddeld_loonsom)

baos_2014 = baos %>% filter(jaar==2014)

gemeente_lonen_2014 = baos_2014 %>% 
  group_by(statcode) %>% mutate(
    totaal_loonsom = sum(loonsom), 
    totaal_onderwijzers = sum(onderwijzers),
    gemiddeld_loonsom = totaal_loonsom / totaal_onderwijzers,
    ) %>% select(
      statcode, totaal_loonsom, totaal_onderwijzers, gemiddeld_loonsom, loc_randstad
    )

summary(lm(data=gemeente_lonen_2014, gemiddeld_loonsom~loc_randstad))

boxplot(gemiddeld_loonsom~loc_randstad, data=gemeente_lonen_2014, xlab= "Randstad", ylab="Gemiddeld Loon")
hist(gemeente_lonen_2014$gemiddeld_loonsom)
  
```

```{r}
library(sf)
gemeentegrenzen <- st_read("https://geodata.nationaalgeoregister.nl/cbsgebiedsindelingen/wfs?request=GetFeature&service=WFS&version=2.0.0&typeName=cbs_gemeente_2019_gegeneraliseerd&outputFormat=json")

data <-
  gemeentegrenzen %>%
  left_join(gemeente_lonen_2019)

data %>%
  ggplot() +
  geom_sf(aes(fill = gemiddeld_loonsom)) +
  scale_fill_viridis_c(limits=c(2700, 4000)) +
  labs(title = "Gemiddelde Loonsom 2019 (per gemeente)", fill = "") +
  theme_void()

data <-
  gemeentegrenzen %>%
  left_join(gemeente_lonen_2014)

data %>%
  ggplot() +
  geom_sf(aes(fill = gemiddeld_loonsom)) +
  scale_fill_viridis_c(limits=c(2700, 4000)) +
  labs(title = "Gemiddelde Loonsom 2014 (per gemeente)", fill = "") +
  theme_void()

```


```{r}
gemeente_loon_jaar = baos %>% 
  group_by(jaar) %>% mutate(
    jaar_gemiddelde = mean(loonsom)
  ) %>% select(jaar_gemiddelde, jaar)

ggplot(gemeente_loon_jaar, aes(x = jaar, y = jaar_gemiddelde)) + geom_line() + theme_minimal() 


ontwikkeling_loon_gemeente = baos %>%  filter(jaar == 2014 | jaar == 2019) %>% group_by(statcode, brin)  %>%
   summarise(
     dif_loon_school = gem_loonsom[jaar==2019] - gem_loonsom[jaar==2014] 
   ) %>% group_by(statcode) %>%
  summarise(
    gem_gemeente = mean(dif_loon_school)
  )

data <-
  gemeentegrenzen %>%
  left_join(ontwikkeling_loon_gemeente)

data %>%
  ggplot() +
  geom_sf(aes(fill = gem_gemeente)) +
  scale_fill_viridis_c() +
  labs(title = "Verandering Loonsom 2014-2019 (per gemeente)", fill = "") +
  theme_void()

```
```{r}
hist(baos_2019$gem_loonsom)

schaal_11_data = baos_2019 %>% mutate(
  percentage_11 = schaal_11 / onderwijzers
) 

ggplot(schaal_11_data, aes(percentage_11, gem_loonsom)) +
  geom_bin2d()

schaal_11_12_data = baos_2019 %>% mutate(
  percentage_11_12 = (schaal_11 + schaal_12) / onderwijzers
) 

ggplot(schaal_11_12_data, aes(percentage_11_12, gem_loonsom)) +
  geom_bin2d()

```

```{r}
library(readxl)
schoolgewicht <- read_excel("/Users/anne-licnossen/Documents/Min Financiën/R. projects/tekortindicator/schoolweging-2017-2019.xlsx", sheet="Tabel 3") %>%
  select(brin=`Tabel 3`, schoolweging= `...4`) %>% slice(-(1:4)) %>% mutate(schoolweging = as.numeric(schoolweging))

baos_2019_schoolgewicht = baos_2019 %>% left_join(schoolgewicht) %>% mutate(
   percentage_11_12 = (schaal_11 + schaal_12) / onderwijzers
)

ggplot(baos_2019_schoolgewicht, aes(schoolweging, gem_loonsom)) +
  geom_bin2d()

summary(lm(data=baos_2019_schoolgewicht, gem_loonsom~schoolweging+percentage_11_12))


```


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

